% Diploma Project Assignment placeholder. TODO: replace with the official IITU
% Assignment form (the «Задание на дипломный проект» sheet your department issues).

\thispagestyle{empty}

\begin{center}
{\large MINISTRY OF SCIENCE AND HIGHER EDUCATION OF THE REPUBLIC OF KAZAKHSTAN\par}
\vspace{0.2cm}
{\large \textbf{INTERNATIONAL INFORMATION TECHNOLOGY UNIVERSITY}\par}
\vspace{0.5cm}
\begin{flushright}
``APPROVED'' \\
Head of Department TODO \\
\rule{4cm}{0.4pt}~/ TODO / \\
``\rule{1cm}{0.4pt}'' \rule{2cm}{0.4pt} 2026
\end{flushright}

\vspace{0.5cm}
{\Large \textbf{ASSIGNMENT}\par}
{\large for the diploma project of the student\par}
\vspace{0.3cm}
{\large \textbf{TODO: Student full name}\par}
\end{center}

\vspace{0.5cm}

\noindent\textbf{1. Topic of the project:} ICS Anomaly Detection for Cyber-Physical Security.\\[0.2cm]
Approved by the order of the rector No.~TODO from ``\rule{1cm}{0.4pt}'' \rule{2cm}{0.4pt} 2026.

\vspace{0.3cm}
\noindent\textbf{2. Deadline for submission:} ``\rule{1cm}{0.4pt}'' \rule{2cm}{0.4pt} 2026.

\vspace{0.3cm}
\noindent\textbf{3. Initial data:} HAI 21.03 testbed dataset (Hyundai/HanyangIIoT) and Morris gas-pipeline dataset (Mississippi State University); six anomaly-detection models from the literature; PyTorch / scikit-learn implementation environment.

\vspace{0.3cm}
\noindent\textbf{4. List of issues to be developed:}
\begin{enumerate}
  \item Survey of ICS anomaly detection methods and evaluation metrics (literature review).
  \item Implementation of six anomaly detectors (Isolation Forest, OC-SVM, Dense Autoencoder, LSTM Autoencoder, USAD, TranAD) under a unified interface.
  \item Implementation of three time-aware evaluation metrics (point-wise, point-adjust, eTaPR) with unit tests.
  \item Cross-testbed transfer study: HAI~$\leftrightarrow$~Morris under source- and target-calibrated thresholds; within-HAI leave-one-process-out control.
  \item Per-sensor attribution evaluation against process-level attack labels.
  \item Discussion of findings and practical implications for deployment.
\end{enumerate}

\vspace{0.3cm}
\noindent\textbf{5. Graphic material:} 5 figures, 8 tables.

\vspace{0.3cm}
\noindent\textbf{6. Recommended literature:} Tuli et al. 2022 (TranAD); Audibert et al. 2020 (USAD); Shin et al. 2020 (HAI dataset); Morris \& Gao 2014 (Morris dataset); Kim et al. 2022 (PA-F1 critique); Hwang et al. 2022 (eTaPR).

\vspace{0.5cm}
\noindent\textbf{Date of issue:} ``\rule{1cm}{0.4pt}'' \rule{2cm}{0.4pt} 2026.

\vspace{1cm}
\noindent
\begin{tabular}{p{6cm}p{4cm}}
Scientific advisor & \rule{4cm}{0.4pt} \\
TODO: Full name, degree & (signature) \\
\\
Student & \rule{4cm}{0.4pt} \\
TODO: Full name & (signature) \\
\end{tabular}

\newpage
