\chapter{Introduction}
\label{ch:intro}

\section{Motivation}

Industrial control systems (ICS) are the interface between digital policy and physical plant: programmable logic controllers, supervisory control and data acquisition (SCADA) servers, and remote terminal units that translate operator intent into actuator commands and report sensor readings back.  Unlike enterprise IT, where a successful intrusion typically results in data theft or service disruption, an undetected intrusion into an ICS can cause damage in the physical world that software-only compromises cannot --- damage to equipment, environmental release, or harm to human operators.  The historical record establishes that this is not a hypothetical concern.  Stuxnet~\cite{langner2013stuxnet} demonstrated that targeted control-loop manipulation can destroy industrial centrifuges while presenting normal-looking telemetry to monitoring stations.  The 2015 Ukraine power-grid attack~\cite{ukraine2016analysis} caused a multi-hour outage affecting roughly 225{,}000 customers across three regional distribution operators, with attackers using legitimate SCADA functionality to open breakers from inside the operators' own control consoles.  Earlier incidents such as the Maroochy Shire wastewater spill (2000) showed that even comparatively unsophisticated insider-style attacks can cause significant environmental harm when basic anomaly-detection telemetry is absent.

The defensive response from the research community has been to construct labelled testbeds and publish anomaly detectors evaluated on them.  Public ICS datasets --- HAI~\cite{shin2020hai}, SWaT~\cite{goh2017swat}, WADI~\cite{ahmed2017wadi}, the Morris gas-pipeline corpus~\cite{morris2014industrial} --- enable reproducible benchmarking, and a large body of deep-learning anomaly detectors reports high F\textsubscript{1} scores on them.  Recent transformer- and adversarial-autoencoder-based detectors (TranAD~\cite{tuli2022tranad}, USAD~\cite{audibert2020usad}) report point-adjusted F\textsubscript{1} above $0.95$ on HAI and SWaT, which would suggest a near-solved problem.  An operator considering deployment, however, faces three questions that the headline numbers skip over:

\begin{enumerate}
  \item \textbf{Do the rankings survive time-aware evaluation?}  Point-wise F\textsubscript{1} treats every timestep as independent.  Point-adjust F\textsubscript{1} (PA-F\textsubscript{1})~\cite{xu2018unsupervised} inflates scores by granting full credit when any single point inside an attack event is flagged --- a recent critique~\cite{kim2022towards} shows that under PA-F\textsubscript{1} a random detector can outperform genuine ones.  eTaPR~\cite{hwang2022etapr} attempts to restore event-level rigour by weighting detection coverage and timeliness.  Which metric one reports changes which model ``wins.''
  \item \textbf{Do these detectors generalize across testbeds?}  All cited papers train and test on the same dataset.  An HAI-trained model should, in principle, capture ICS-general structure and transfer to an industrial gas-pipeline testbed --- or the claim of ``anomaly detection'' is overstated.  A detector that only works on the testbed it was trained on is, in operational terms, a one-off model fitted to a particular plant.
  \item \textbf{Can the detector localize?}  A flagged window is actionable only if an operator can see \emph{which sensor} is implicated.  Detection without localization forces a manual triage step that scales poorly when alerts fire across hundreds of sensors.  We evaluate per-feature attribution against HAI's per-process attack labels and ask whether the best-detecting model is also the best-localizing one.
\end{enumerate}

\section{Problem statement}
\label{sec:problem}

The state of the published literature can be summarized as follows.  Detection performance is reported as a single F\textsubscript{1}-style number (most often PA-F\textsubscript{1}) on a single testbed under a single threshold-selection protocol.  This is enough to compare new architectures against prior architectures \emph{within the same evaluation envelope}, but it is not enough to support an operational claim that a published detector will work on a new plant under a metric that an operator actually cares about.  The gap between published F\textsubscript{1} and deployable performance is widely acknowledged in conversation but, to our knowledge, has not been quantified end-to-end on a common set of detectors and testbeds in a way that lets a reader read off how much of the published number is real and how much is an artifact of the evaluation envelope.

This thesis quantifies that gap.  The thesis treats existing detectors, datasets, and metrics as given and constructs an evaluation grid that varies (a)~the scoring metric, (b)~the calibration regime, (c)~the testbed used for training versus testing, and (d)~the evaluation question (detection versus per-sensor localization).  The intent is not to crown a winner but to expose where rankings are protocol-stable and where they are protocol-artifacts, and to provide a reproducible apparatus for future work to extend.

\section{Research objectives}
\label{sec:objectives}

The work is structured around five concrete objectives, each of which is realized by a specific component of the open-source codebase released alongside the thesis:

\begin{enumerate}
  \item \textbf{Implement six anomaly detectors under a unified interface} --- two classical baselines (Isolation Forest, One-Class SVM), two autoencoder baselines (Dense, LSTM), and two state-of-the-art detectors (USAD, TranAD) --- such that swapping models in an experiment requires only a configuration change.
  \item \textbf{Implement three time-aware evaluation metrics} (point-wise, point-adjust, eTaPR) with hand-computed unit tests, so that ranking changes between metrics are demonstrably attributable to the metrics themselves and not to implementation differences.
  \item \textbf{Run a same-dataset detection benchmark} for all six models on both HAI~21.03 and Morris under all three metrics, and document any reproducibility gaps against the original papers honestly rather than tuning until they close.
  \item \textbf{Run a cross-testbed transfer study} with a feature-type-vector schema alignment, under both source-calibrated and target-calibrated thresholds, and a within-HAI leave-one-process-out (LOPO) control that isolates feature-distribution shift from class-prior shift.
  \item \textbf{Run a per-sensor attribution study} that evaluates each model's per-feature attribution against HAI's per-process attack labels, including a multi-seed sensitivity check, to test whether the best-detecting models are also the best-localizing models.
\end{enumerate}

\section{Contributions}

\paragraph{C1 --- Cross-testbed transfer protocol that separates representation from threshold.}  We project HAI (79 features) and Morris (16 features) onto a shared 6-dim type-vector (pressure, pump\_state, setpoint, system\_state, valve\_position, control\_signal) and run transfer under two calibration regimes: source-domain percentile (\code{src-cal}) and target-normal percentile (\code{tgt-cal}).  We show that \code{src-cal} collapses every detector to the predict-all trivial baseline of the \emph{target} class prior --- a known artifact but, to our knowledge, unquantified in the ICS transfer literature.  Under \code{tgt-cal}, windowed SOTA models (LSTM-AE, USAD, TranAD) achieve $0.38$--$0.44$ eTaPR on HAI$\rightarrow$Morris, the first clear classical-vs-SOTA separation in our study.

\paragraph{C2 --- Within-HAI leave-one-process-out (LOPO) experiment isolating feature shift.}  Holding the class prior fixed, we drop one process's features (P1, P2, P3, or P4) and measure per-attacked-process F\textsubscript{1}.  Dropping P2 (turbine controller, discrete state flags) causes \emph{targeted} blindness to P2-targeting attacks (Dense AE pointwise F\textsubscript{1} drops from $0.41$ on P1-attacks to $0.03$ on P2-attacks); dropping P1 (water loop, $82\%$ of test attacks) causes \emph{global} degradation.  This sets a within-testbed ceiling on what sensor-level transfer can achieve, and contextualizes the lower cross-testbed numbers.

\paragraph{C3 --- Process-level attribution evaluation with a detection-vs-localization tradeoff.}  Every model exposes \code{attribute(X)}$\rightarrow$\code{(N, F)}; we evaluate precision@k of the top-k attributed features against HAI's per-process attack labels.  Dense AE --- the weakest Phase-2 detector --- is the only model whose attribution beats the random baseline on all three attacked processes (lift $1.4\times$, $2.7\times$, $3.7\times$ at p@5, averaged over $3$ seeds).  The windowed SOTA models concentrate attribution on P1 features regardless of target, and attention-weighted rollout for TranAD produces numerically identical numbers to plain reconstruction attribution --- a null finding that questions the usefulness of attention-XAI for per-sensor ICS localization.

\section{Practical relevance}
\label{sec:relevance}

The findings have direct operational consequences.  An operator who deploys a published detector trained on HAI to a new plant, and uses the source-domain percentile threshold from the publication, will, by our measurement, get a classifier whose F\textsubscript{1} reflects the new plant's class prior rather than the detector's actual discrimination ability.  An operator who picks ``the best'' detector by published PA-F\textsubscript{1} and expects per-sensor localization for free will, by our measurement, get the worst-localizing model in the comparison.  Both failure modes are silent under standard evaluation; both are visible once the evaluation grid is widened.  A central recommendation of this thesis is therefore methodological rather than architectural: practitioners and reviewers should ask for the full grid before treating a single F\textsubscript{1} as evidence of operational fitness.

\section{Structure of the thesis}

Chapter~\ref{ch:background} surveys the datasets, detectors, and metrics used in the study, including the published critiques of point-adjust F\textsubscript{1} that motivate our triple-metric reporting.  Chapter~\ref{ch:method} describes the implementation: unified \code{AnomalyDetector} interface, scaler-leak guards, schema-align projection, LOPO helpers, attribution evaluation.  Chapter~\ref{ch:results-detection} presents the same-dataset detection results (the metric-sensitivity component).  Chapter~\ref{ch:results-transfer} is the cross-testbed and LOPO study, the headline chapter of the thesis.  Chapter~\ref{ch:results-attribution} is the attribution study and reports the detection-versus-localization tradeoff.  Chapter~\ref{ch:discussion} addresses reviewer-anticipated objections and limitations.  Chapter~\ref{ch:conclusion} summarizes the research outcomes.

\section{Scope and non-goals}

\begin{itemize}
  \item We do not propose a new detector architecture.  The argument is empirical: existing detectors evaluated under a protocol that makes their brittleness visible.
  \item We do not use HAI 22.04 or 23.05 despite their larger attack sets.  HAI 21.03 was already available via git-standard clone and has per-process attack tags sufficient for the attribution study.  HAI 22.04 with per-sensor attack labels is listed as follow-up.
  \item We do not evaluate against SWaT or WADI.  Those are single-testbed benchmarks like the ones we already use; adding them does not change the argument of the cross-testbed contribution.
  \item Defensive deployment and real-time performance are out of scope.  Every reported number is from offline batched evaluation.
  \item We do not evaluate adversarial robustness of the detectors against an attacker who knows the model.  Adversarial ML in the ICS setting is a substantial research area in its own right and is left for future work.
\end{itemize}
