\chapter{Conclusion}
\label{ch:conclusion}

\section{Summary of the work}

This thesis benchmarked six ICS anomaly detectors --- three classical (Isolation Forest, One-Class SVM, Dense Autoencoder) and three windowed state-of-the-art models (LSTM Autoencoder, USAD, TranAD) --- on HAI~21.03~\cite{shin2020hai} and the Morris gas-pipeline dataset~\cite{morris2014industrial} under three time-aware metrics, evaluated cross-testbed transfer under two calibration regimes with a within-HAI leave-one-process-out control, and evaluated per-sensor attribution at the process level using a multi-seed sensitivity check.  All numerical results are produced by configuration-driven runs; every figure and every table cell is reproducible from a committed YAML configuration via the \file{experiments/} command-line interface.  Tests covering the scoring metrics, the data-loading leak guards, and a smoke test for every model class accompany the implementation.

\section{Research outcomes}

Three findings define the thesis:

\begin{enumerate}
  \item \textbf{Source-calibrated cross-testbed transfer is a class-prior artifact.}  Under the common practice of fixing the threshold from source-validation data, every detector degenerates to the predict-all trivial baseline of the target's class prior --- F\textsubscript{1}~$\approx 0.95$ on Morris (whose post-windowing attack rate is $91\%$) and F\textsubscript{1}~$\approx 0.05$ on HAI ($2.4\%$ attack rate).  Target-calibration reveals a meaningful classical-vs-SOTA separation invisible to the source-calibrated protocol: the windowed SOTA models reach $0.38$--$0.44$ eTaPR on HAI$\rightarrow$Morris, whereas the classical baselines do not exceed $0.13$ (\S\ref{sec:finding-src-cal}--\S\ref{sec:finding-tgt-cal}).

  \item \textbf{Feature shift produces targeted, not global, detection blindness.}  Within-HAI LOPO shows that dropping the turbine-controller features specifically blinds detectors to turbine-targeted attacks (Dense Autoencoder pointwise F\textsubscript{1} drops from $0.41$ to $0.03$ on the affected process); dropping the water-loop features globally degrades detection because they carry the bulk of the ``normal'' signal.  Sensor-level transfer within one testbed tops out at approximately $0.57$ pointwise F\textsubscript{1} --- an order of magnitude better than cross-testbed transfer, a useful upper bound on what schema-level alignment can plausibly achieve (\S\ref{sec:finding-lopo}).

  \item \textbf{Detection and localization are a genuine tradeoff.}  The weakest detector by F\textsubscript{1} (Dense Autoencoder) is the only model whose per-feature attribution beats a random baseline on all three attacked processes (precision@5 lift of $1.4\times$, $2.7\times$, and $3.7\times$).  Windowed SOTA models concentrate attribution on the majority-attacked process regardless of the actual target, and attention-weighted rollout for TranAD provides no additional localization signal over plain reconstruction-error attribution (\S\ref{sec:finding-attn-null}).  For an operator triaging a flagged window, the Dense Autoencoder's per-feature output is the more actionable one even though its detection F\textsubscript{1} is lower.
\end{enumerate}

\section{Practical recommendations}

The findings translate into three operational recommendations for practitioners deploying ICS anomaly detectors and for reviewers evaluating new detector publications:

\begin{itemize}
  \item \textbf{Calibrate the decision threshold on target-domain normal data, not on source-domain validation data.}  Any production deployment should collect a clean-operation reference period from the target plant before activating the detector.  A source-trained percentile threshold transferred unchanged produces a trivial classifier whose apparent F\textsubscript{1} reflects nothing more than the target's class prior.
  \item \textbf{Report multiple metrics and treat single-metric rankings with suspicion.}  Pointwise F\textsubscript{1} for per-timestep claims, PA-F\textsubscript{1} only when ``at least one timestep flagged'' is the operationally meaningful event (and with the inflation explicitly acknowledged~\cite{kim2022towards}), eTaPR~\cite{hwang2022etapr} for event-level detection with coverage weighting.
  \item \textbf{Audit attribution alongside detection.}  Where sensor-level localization matters operationally, evaluate the attribution mechanism separately from detection F\textsubscript{1}.  A hybrid pipeline using a SOTA model for flagging and a Dense Autoencoder for sensor attribution may outperform either model alone.
\end{itemize}

\section{Methodological contribution}

Beyond the empirical findings, the thesis contributes a reproducible benchmarking protocol and an open-source implementation that future work can extend without re-deriving the apparatus from scratch.  The unified \code{AnomalyDetector} interface in \file{src/models/base.py}, the leak-guarded data-loading pipeline in \file{src/data\_loader.py} and \file{src/preprocessing.py}, and the configuration-driven experiment runners in \file{experiments/} together constitute infrastructure that lowers the marginal cost of every additional model, dataset, or metric study.  Adding a seventh detector amounts to writing one Python class implementing the abstract base; adding a third dataset amounts to writing one loader function returning the unified \code{(timestamps, features\_df, labels, attack\_ids)} schema.

\section{Outlook}

Future work (discussed in Chapter~\ref{ch:discussion}, \S\ref{sec:future}) includes per-sensor attribution on HAI~22.04, where the published per-attack target-sensor descriptions can be parsed into a ground-truth mapping that would upgrade the localization claim from process level to sensor level; multi-seed attribution across the full six-model set to confirm the LSTM-AE~$\equiv$~USAD attribution coincidence reported in Chapter~\ref{ch:results-attribution}; and a learned-projection variant of the schema alignment to test whether the cross-testbed ceiling moves under a richer alignment than the per-variable type vector used here.

The single methodological recommendation this thesis makes to the ICS anomaly detection community is to report the full grid --- multiple metrics, multiple calibration regimes, attribution alongside detection --- rather than the single headline F\textsubscript{1} that currently dominates publications.  Every number in this document is reproducible by running \code{python -m experiments.run[\_transfer|\_lopo|\_attribution] <config.yaml>}.  We hope the reproducibility apparatus is at least as useful to subsequent work as the specific findings themselves.
